---
title: "R Notebook"
output: html_notebook
---

Here, we begin the analysis on RNAse +/- control experiment from E.coli (page 164 from notebook 3 for further info)

Experimental info. 5 samples in the stationary phase, (in LBM9?) treated with 700 mJ/cm2 in 15cm plate. For each sample there is an RNAse + and - run. Also- need to add detail of the MS run, note that these were analyzed independently (i.e. no match between runs). From the samples there were 2 (1 per group) that had to discarded from the analysis. The chromatograms were really bad. All the others were fine! 


```{r}
#### Load packages ####
library(tidyverse)
library(MSnbase)
library(GO.db)
library(biobroom)
library(camprotR)
library(plyr)
library(dplyr)
library(naniar)

select <- dplyr::select

suppressMessages(source("../../ecoli-rnase-main/shared_files/Utility.R", chdir = TRUE)) # need to change the path
suppressMessages(source("../../ecoli-rnase-main/shared_files/OOPS_analysis.R", chdir = TRUE))
```

```{r}
# set up standardized plotting scheme
theme_set(theme_bw(base_size = 20) +
           theme(panel.grid.major=element_blank(),
                 panel.grid.minor=element_blank(),
                 aspect.ratio=1, 
                 plot.title = element_text(hjust = 0.5)))
```

Read individual files into a master df
```{r}
infiles <- Sys.glob('../raw/SeparateSearches/*PeptideGroups.txt')[1:2]
df <- read.csv("../analysos/sample_names.csv") %>% mutate( concated_column = paste0(TRT, "_", rep, "_"))

allData <- lapply(infiles, function(.file){
  # read in file
  .file = infiles[1]
  dat <- read.delim(.file, header=T)
  
  for(i in 1:length(colnames(dat))){
    var1 <- colnames(dat)[i]
    for(j in 20:1){
      #print('we are now processing')
      #print(j)
      samplen <- df$samples[j]
      if(grepl(samplen, var1)==TRUE){
       colnames(dat)[i] <- gsub(samplen, df$concated_column[j], colnames(dat)[i])
        # extract important terms
      }}}
    file_name <- strsplit(.file, "_")[[1]]
    # add the df
    dat$id <- paste0(file_name[4], "_", file_name[3])
    colnames(dat)
})
```

```{r}
identical(names(allData[[1]]), names(allData[[2]]) )
# combine into a single dataframe
myDF <- do.call(rbind, allData)
print(table(myDF$id))

test <- duplicated(myDF)
table(test)

#tiff('../figures/peptide_count_per_sample.tiff', units="in", width=5, height=4, res=300, compression = 'lzw')
ggplot(data=as.data.frame(table(myDF$id)), aes(x=sub(".", "", Var1), y=Freq)) + geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + xlab("") + ylab("peptide count")
#dev.off()
```
get list of crap accessions for filter step 
```{r}
# Load the cRAP FASTA used for the PD search
crap.fasta <- Biostrings::fasta.index(
  "../cRAP_FullIdentifiers.fasta",
  seqtype = "AA"
)

# see species plot
spcvec = c()
for(i in 1:length(crap.fasta$desc)){
  tmp = strsplit(crap.fasta$desc[i], "OS=")
  tmp = strsplit(tmp[[1]][2], " ")
  
  spcvec[i] = paste0(tmp[[1]][1], "_", tmp[[1]][2])
}

print(table(spcvec))

#tiff('../figures/crap_accessions.tiff', units="in", width=5, height=4, res=300, compression = 'lzw')
table(spcvec) %>% as.data.frame() %>% 
  ggplot(aes(y=reorder(spcvec, Freq), x = Freq)) + geom_bar(stat="identity") + ylab("species") +
  ggtitle('crap.fasta by species')
#dev.off()

# Extract the non cRAP UniProt accessions associated with each cRAP protein
crap.accessions <- crap.fasta %>% 
  pull(desc) %>% 
  stringr::str_extract_all("(?<=\\|).*?(?=\\|)") %>% 
  unlist() %>% 
  .[!. %in% "000000"] # remove invalid UniProt accessions
``` 
this crap list has a unique ecoli protein... need to look into securing a more comprehensive list.

For now, lets use the functions in camprotR to parse the infile
```{r}
.data <- camprotR::parse_features(myDF, 
                                  filter_crap = TRUE, 
                                  crap_proteins = crap.accessions, filter_associated_crap = TRUE, 
                                  unique_master = FALSE)
# again, only 1 protein supplied as crap protein. 
# add extra filters for contaminants


length(colnames(myDF))
length(unique(colnames(myDF)))


.data <- .data %>% filter(Contaminant=='False')
camprotR:::message_parse(.data, 'Master.Protein.Accessions', 'removed remaining contaminants')

.data <- .data %>% filter(Quan.Info!='NoQuanValues')
camprotR:::message_parse(.data, 'Master.Protein.Accessions', 'removed remaining contaminants')

```

Now, lets subset to the required abundance columns. This has not been analyzed by MBR (all samples analyzed independently)
```{r}
all_abun_cols <- colnames(.data)[grepl('^Abundance', colnames(.data))]
print(all_abun_cols)

abundance <- .data %>% 
  select(c(Sequence, Modifications, Number.of.Missed.Cleavages,
           Master.Protein.Accessions, Master.Protein.Descriptions, Abundance.F.Sample.B, id)) %>%
  tidyr::pivot_wider(names_from=id, values_from=Abundance.F.Sample.B)

print(head(abundance))

# remove peptides with all missing vals
abundance <- abundance %>% mutate(missingVals = rowSums(is.na(abundance)))
print(table(abundance$missingVals))

abundance <- abundance %>% filter(missingVals != 8)
print(table(abundance$missingVals))
```


store the data into an msnset to manipulate the peptide object into protein vals
```{r}
eDat <- as.matrix(abundance[grepl('^B', colnames(abundance))])
fDat <- abundance[,1:5] %>% as.data.frame()

peptide_res <- MSnSet(exprs=eDat, fData=fDat)

rownames(peptide_res) <- paste(fData(peptide_res)$Sequence, 
                               fData(peptide_res)$Modifications, 
                              # fData(peptide_res)$Number.of.Missed.Cleavages,
                               fData(peptide_res)$Master.Protein.Accessions,
                               sep=":")

pDat <- pData(peptide_res) %>%
  mutate(rep = gsub('[[:alpha:]]+_', '', rownames(.)),
         sample = sub(".", "", gsub('_[[:digit:]]+', '', rownames(.))))

# add the pData using the sample names (in the right order)
pData(peptide_res) <- pDat[colnames(peptide_res),]
```

Let's check the sum and max intensity per sample
```{r}
plot_data <- exprs(peptide_res) %>% melt() %>% 
  mutate(rep =  gsub('[[:alpha:]]+_', '', Var2), 
         type = sub(".", "", gsub('_[[:digit:]]+', '', Var2))) %>%
  group_by(rep, type) %>% 
  dplyr::summarise(sum_intensity=sum(value, na.rm=TRUE),
              max_intensity=max(value, na.rm=TRUE),
              median_intensity=median(value, na.rm=TRUE),
              length_intensity=sum(is.finite(value)))

p <- plot_data %>% 
  ggplot(aes(type, sum_intensity, fill=type, group=type)) +
    geom_bar(stat='identity', position='dodge') +
    theme_bw() +
    facet_wrap(~rep)

#tiff('../figures/sum_intensity.tiff', units="in", width=5, height=4, res=300, compression = 'lzw')
print(p + ggtitle("sum intensity"))
#dev.off()

#tiff('../figures/max_intensity.tiff', units="in", width=5, height=4, res=300, compression = 'lzw')
print(p + aes(y=max_intensity) + ggtitle("max intensity"))
#dev.off()

#tiff('../figures/meidan_intensity.tiff', units="in", width=5, height=4, res=300, compression = 'lzw')
print(p + aes(y=median_intensity) + ggtitle("median intensity"))
#dev.off()
```
In general, we see a higher intensity in the minus samples. this is expected as rbp's will leave the interface upon the digestion of rna. However, this may also be a bias in the data. the rnase+ samples are harder to process (the rna is what makes the pellet visible).

```{r}
#tiff('../figures/peptide_quant_log10_boxplot.tiff', units="in", width=5, height=4, res=300, compression = 'lzw')
plot_quant(log(peptide_res, base=10), method='box') +
  scale_x_discrete(labels=rownames(pData(peptide_res)))
#dev.off()

#tiff('../figures/peptide_quant_log10_density.tiff', units="in", width=5, height=4, res=300, compression = 'lzw')
plot_quant(log(peptide_res, base=10), method='density') +
  scale_colour_manual(labels=rownames(pData(peptide_res)),
                      values=rep(get_cat_palette(4), 3))
#dev.off()
```

Next we want to remove peptides which are not seen in enough samples and proteins which have too few peptides. Below we test a range of values for 'min_samples' and 'min_peptides' to see how this affects the number of peptides and proteins we retain.
Note that at this point we still have lipoproteins since these are required downstream for the normalization. The protein count is excluding lipoproteins (which are shown separately).

```{r}
CheckFilteringOptions <- function(peptide_res,
                                  min_samples_range=1:8,
                                  min_peptides_range=1:4){
  rows <- NULL
  for(min_samples in min_samples_range){
    for(min_peptides in min_peptides_range){
      print(c(min_samples, min_peptides))
      .data <- restrict_features_per_protein(
        peptide_res, min_features=min_peptides, plot=FALSE)
      
      .data <- filterNA(.data, pNA=1-(min_samples/ncol(.data)))
      
      .data <- restrict_features_per_protein(
        peptide_res, min_features=min_peptides, plot=FALSE)
      
      .data <- filterNA(.data, pNA=1-(min_samples/ncol(.data)))
      
      rows[[paste0(min_samples, min_peptides)]] <- fData(.data) %>%
        select(Sequence, Modifications, Master.Protein.Accessions) %>%
        mutate('min_samples'=min_samples, 'min_peptides'=min_peptides)
    }
  }
  
  peptide_filtering_options <- data.frame(do.call('rbind', rows))
  
  return(peptide_filtering_options)
}

peptide_filtering_options <- CheckFilteringOptions(peptide_res)
```

```{r}
p <- peptide_filtering_options %>%
  group_by(min_peptides, min_samples) %>%
  tally() %>%
  ggplot(aes(min_samples, n,
             colour=factor(min_peptides),
             group=min_peptides)) +
  geom_line() +
  geom_point() +
  scale_colour_manual(values=get_cat_palette(4),
                      name='Minimum peptides\nper protein') +
  theme_camprot(base_size = 15) +
  ylim(0, NA) +
  xlab('Minimum samples with peptide detected') +
  ylab('Peptides retained') +
  scale_x_continuous(breaks=seq(2,12,2))
  
#tiff('../figures/peptide_retained_post_filter.tiff', units="in", width=5, height=4, res=300, compression = 'lzw')
print(p)
#dev.off()
```
lets filter at `min_samples` = 2 and `min_peptides` = 2. This is very relaxed, but we are going to validate candidates in the lab.

```{r}
filterPeptides <- function(peptide_abundance, min_samples, min_peptides){
  .data <- restrict_features_per_protein(
        peptide_abundance, min_features=min_peptides, plot=FALSE)
      
  .data <- filterNA(.data, pNA=1-(min_samples/ncol(.data)))
  
  .data <- restrict_features_per_protein(
        peptide_abundance, min_features=min_peptides, plot=FALSE)
      
  .data <- filterNA(.data, pNA=1-(min_samples/ncol(.data)))
  
  .data <- restrict_features_per_protein(
        peptide_abundance, min_features=min_peptides, plot=FALSE)
      
  .data <- filterNA(.data, pNA=1-(min_samples/ncol(.data)))
  
  return(.data)
}
```

```{r}
peptide_abundance_filtered <- filterPeptides(peptide_res,
                                             min_samples=3,
                                             min_peptides=2)

```


Later, we will want to return the question of how the min_samples and min_peptides affects the results so below we create an object to help us track this
```{r}
colnames(pData(peptide_abundance_filtered))[2] <- "sample_name"

getPeptideProteinStats <- function(peptide_abundance_filtered){
  peptide_abundance_long <- biobroom::tidy.MSnSet(peptide_abundance_filtered, addPheno=TRUE) %>%
    filter(is.finite(value)) %>%
    merge(fData(peptide_abundance_filtered), by.x='protein', by.y='row.names')
  
  n_samples_per_pep <- peptide_abundance_long %>%
    group_by(protein) %>% tally()
  
  n_pep_per_protein <- peptide_abundance_long  %>%
    group_by(Master.Protein.Accessions, sample_name) %>%
    tally(name='n_peptides')
  
  mean_samples_per_pep_per_protein <- peptide_abundance_long  %>%
    merge(n_samples_per_pep, by='protein') %>%
    group_by(Master.Protein.Accessions, sample_name) %>%
    dplyr::summarise('n_samples'=mean(n))
  
  peptide_protein_stats <- mean_samples_per_pep_per_protein %>%
    merge(n_pep_per_protein, by=c('Master.Protein.Accessions', 'sample_name'))
  return(peptide_protein_stats)
  
}

peptide_protein_stats <- getPeptideProteinStats(peptide_abundance_filtered)
```

Below, we aggregate to protein level abundance using the `robust` method in MSnbase. At the same time, we store information in the feature data about the number of peptides used to quantify the protein.

```{r}
protein_abundance <- log(peptide_abundance_filtered, base=2) %>%
  combineFeatures(method='robust',
                  groupBy=fData(peptide_abundance_filtered)$Master.Protein.Accessions,
                  maxit=1000L)
                              
peptides <- count_features_per_protein(peptide_abundance_filtered) %>%
  pivot_wider(names_from=sample, values_from=n) %>%
  tibble::column_to_rownames('Master.Protein.Accessions')
fData(protein_abundance)$peptides <- peptides[,colnames(protein_abundance)]
```

Below, we summarize the number of missing values at peptide, filtered peptides and protein level
```{r}
#tiff('../figures/filter22Unpaired/plot_na_peptide.tiff', units="in", width=10, height=8, res=300, compression = 'lzw')
plotNA(peptide_res)
#dev.off()

#tiff('../figures/filter22Unpaired/plot_na_protein.tiff', units="in", width=10, height=8, res=300, compression = 'lzw')
plotNA(protein_abundance)
#dev.off()
```
```{r, fig.height=10, fig.width=15}
#library(naniar)
dim(peptide_res)
table(rowSums(is.na(exprs(peptide_res))))

#tiff('../figures/filter22Unpaired/upset_na_peptide.tiff', units="in", width=12, height=8, res=300, compression = 'lzw')
gg_miss_upset(data.frame(exprs(peptide_res)), nsets=12,
              nintersects=20, keep.order=TRUE)
#dev.off()
dim(protein_abundance)
table(rowSums(is.na(exprs(protein_abundance))))
#tiff('../figures/filter22Unpaired/upset_na_protein.tiff', units="in", width=12, height=8, res=300, compression = 'lzw')
gg_miss_upset(data.frame(exprs(protein_abundance)), nsets=12,
              nintersects=20, keep.order=TRUE)
#dev.off()
```

```{r}
x <- protein_abundance[,pData(protein_abundance)$sample_name=='minus']
table(rowSums(is.na(exprs(x))))
#tiff('../figures/filter22Unpaired/upset_na_protein_minus_samples.tiff', units="in", width=5, height=4, res=300, compression = 'lzw')
gg_miss_upset(data.frame(exprs(x)), nsets=12,
              nintersects=20, keep.order=TRUE)
#dev.off()
x <- protein_abundance[,pData(protein_abundance)$sample_name!='minus']
table(rowSums(is.na(exprs(x))))
#tiff('../figures/filter22Unpaired/upset_na_protein_plus_samples.tiff', units="in", width=12, height=8, res=300, compression = 'lzw')
gg_miss_upset(data.frame(exprs(x)), nsets=12,
              nintersects=20, keep.order=TRUE)
#dev.off()
```
Save out the objects to use in subsequent notebooks


```{r}
#saveRDS(peptide_res, '../results/peptide_abundance.rds')
#saveRDS(peptide_abundance_filtered, '../results/peptide_abundance_filtered_22.rds')
#saveRDS(protein_abundance, '../results/protein_abundance.rds')
saveRDS(peptide_protein_stats, '../results/peptide_protein_stats.rds')
```
